\begin{acknowledgements}
	Firstly, I would like to thank my supervisors, Tobias Golling and Slava Voloshynovskiy.
	They have been incredibly supportive over the past four years, providing guidance while also encouraging me to explore my ideas.

	I want to thank my other collaborators at both the University of Geneva and ATLAS, with special mention to Knut Zoch, Manuel Guth, Johnny Raine, Micheal Kagan, and Lukas Heinrich.

	I want to thank my friends at the office in Ecole de Medecine who have made the past four years not just bearable but, dare I say, somewhat enjoyable.

	I would like to also thank my proofreaders for this thesis.
	First is my mother, who diligently found the thousands of errors in this thesis, some of which I have corrected.
	I live for your inline comments of ``Not English! Rewrite!''.
	My second proofreader is my colleague and friend Debajyoti Sengupta, who finished his own thesis a few months before me.
	I didn't help him with his thesis, but he helped me with mine, and for that, I am grateful.
	I owe you many nights of dinner and wine.

	Finally, I would like to Thejal Mathura, who started this journey with me as my partner, became my fiancé, and is now my wife.
	Thank you for your unwavering belief in me, endless support when I was down, and infinite patience in listening to me explain score-based diffusion models while we were on holiday.
	I love you.
\end{acknowledgements}

\clearpage

\thispagestyle{empty}
\null\vfill

\begin{center}

	\begin{minipage}{0.7\textwidth}
		\raggedright{\large\itshape%
			The most important step a man can take. \\
			It's not the first one, is it? \\
			It's the next one. \\
			Always the next step... \\\par\bigskip
		}
		\raggedleft\large\MakeUppercase{Brandon Sanderson}\par%
	\end{minipage}
\end{center}

\vfill\vfill

\clearpage
